Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
Edits to Discussion
Edits to Discussion cont.

 % Table 4 Assigment to maize and teosinte in tb1 region and chr 1 
\begin{table}[htbp]
\rowcolors{2}{gray!25}{white}
  \centering
  \caption{Assignments to maize and teosinte in the \emph{tb1} and chromosome 1 regions from STRUCTURE}
    \begin{tabular}{ccccc} \
          & \multicolumn{2}{c}{\textbf{\emph{tb1} region}} & \multicolumn{2}{c}{\textbf{Chr 1}} \\\toprule
    \bf{Population} & \bf{Maize} & \bf{Teosinte} & \bf{Maize} & \bf{Teosinte} \\\midrule
    Ejutla A & 0.022 & 0.978 & 0.203 & 0.797 \\
    Ejutla B & 0.019 & 0.981 & 0.187 & 0.813 \\
    La Mesa & 0.012 & 0.988 & 0.193 & 0.807 \\
    San Lorenzo & 0.016 & 0.984 & 0.205 & 0.795 \\\bottomrule
    \end{tabular}
  \label{Table4Q}
\end{table}

@@

Adaptation occurs either due to selection on standing variation or on \emph{de novo} mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems; for example, selection for lactose tolerance in humans \citep{Plantinga2012, Tishkoff2007}, variation at the \emph{Eda} locus in three-spined stickleback \citep{Kitano2008, Colosimo2005}, and pupal diapause in the Apple Maggot fly \citep{Feder2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (\emph{RAMOSA1}, \citealt{SigmonVollbrecht2010}; \emph{barren stalk1}, \citealt{Gallavotti2004}; and \emph{grassy tillers1}, \citealt{Whipple2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), providing the first evidence that the \emph{Hopscotch} insertion is segregating in teosinte \citep{Studer2011}.

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with \citet{Studer2011} in that we identified the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus in this region than in the Balsas River Basin where maize was domesticated \citep{Matsuoka2002}. Potential explanations for the high frequency of the \emph{Hopscotch} element in \emph{parviglumis} from Jalisco include gene flow from maize, genetic drift, and natural selection.

While gene flow from crops into their wild relatives is well-known, \citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}, our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize; \citealt{Hufford2013}). We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (Fig. \ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in some of  our \emph{parviglumis} populations.

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the \emph{Hopscotch} in these populations. If the \emph{Hopscotch} allele was introgressed in the distant past, recombination may have broken up LD, a pattern that would be consistent with our data.  We find this scenario less plausible, however, as there is no reason why gene flow should have been high in the past but absent in present-day sympatric populations.  In fact, early generation maize-teosinte hybrids are easy to find in these populations today (MB Hufford, pers. observation), and genetic data support ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, VanHeerwaarden2011}.

>>>>

 % Table 4 Assignment to maize and teosinte in tb1 region and chr 1 
\begin{table}[htbp]
\rowcolors{2}{gray!25}{white}
  \centering
  \caption{Assignments to maize and teosinte in the \emph{tb1} and chromosome 1 regions from STRUCTURE}
    \begin{tabular}{ccccc} \
          & \multicolumn{2}{c}{\textbf{\emph{tb1} region}} & \multicolumn{2}{c}{\textbf{Chr 1}} \\\toprule
    \bf{Population} & \bf{Maize} & \bf{Teosinte} & \bf{Maize} & \bf{Teosinte} \\\midrule
    Ejutla A & 0.022 & 0.978 & 0.203 & 0.797 \\
    Ejutla B & 0.019 & 0.981 & 0.187 & 0.813 \\
    La Mesa & 0.012 & 0.988 & 0.193 & 0.807 \\
    San Lorenzo & 0.016 & 0.984 & 0.205 & 0.795 \\\bottomrule
    \end{tabular}
  \label{Table4Q}
\end{table}

@@

Adaptation occurs due to selection on standing variation or \emph{de novo} mutations. Adaptation from standing variation has been well-described in a number of systems; for example, selection for lactose tolerance in humans \citep{Plantinga2012, Tishkoff2007}, variation at the \emph{Eda} locus in three-spined stickleback \citep{Kitano2008, Colosimo2005}, and pupal diapause in the Apple Maggot fly \citep{Feder2003}. Although the adaptive role of standing variation has been described in many systems, its importance in domestication is not as well studied.

In maize, alleles at domestication loci (\emph{RAMOSA1}, \citealt{SigmonVollbrecht2010}; \emph{barren stalk1}, \citealt{Gallavotti2004}; and \emph{grassy tillers1}, \citealt{Whipple2011}) are thought to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in maize domestication. The \emph{teosinte branched1} gene is one of the best characterized domestication loci, and, while previous studies have suggested that differences in plant architecture between maize and teosinte are a result of selection on standing variation at this locus, little is known about natural variation at this locus and its ecological role in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), providing the first evidence that the \emph{Hopscotch} insertion is segregating in teosinte \citep{Studer2011}.

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, by widely sampling across teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with \citet{Studer2011} in that we identify the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, suggesting a different history of the \emph{tb1} locus in this region than in the Balsas River Basin where maize was domesticated \citep{Matsuoka2002}. Potential explanations for the high frequency of the \emph{Hopscotch} element in \emph{parviglumis} from Jalisco include gene flow from maize, genetic drift, and natural selection.

While gene flow from crops into their wild relatives is well-known, \citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}, our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize). We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (Fig. \ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in a subset of our \emph{parviglumis} populations.

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the \emph{Hopscotch} in these populations. If the \emph{Hopscotch} allele was introgressed in the distant past, recombination may have broken up LD, a process that would be consistent with our data.  We find this scenario less plausible, however, as there is no reason why gene flow should have been high in the past but absent in present-day sympatric populations.  In fact, early generation maize-teosinte hybrids are easy to find in these populations today (MB Hufford, pers. observation), and genetic data support ongoing gene flow between domesticated maize and both \emph{mexicana} and \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, VanHeerwaarden2011}.
